---
title: "Exploratory Data Analysis of SAFI Survey Data"
author: "Short Analytical Report"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## 1. Data description

This report provides a short exploratory data analysis (EDA) of the **SAFI survey dataset**.

**Unit of observation:** each row represents one household.

**Variables:** columns contain household characteristics, location identifiers, food consumption indicators, agricultural information, and related survey attributes.

**Goal of EDA:** quickly check the structure of the data, basic numeric distributions, missingness, and a few simple relationships using `dplyr` for selection/filtering and `ggplot2` for plots.

---

## 2. Data loading

```{r load-packages}
library(dplyr)
library(ggplot2)
library(readr)
```

```{r load-data}
# If SAFI_clean.csv is in the same folder as this .Rmd
safi <- read_csv("data/SAFI_clean.csv")
```

```{r data-structure}
# Data structure: column names, types, and example values
# This helps confirm that the dataset is loaded correctly.
glimpse(safi)
```

**Explanation (table output):** `glimpse()` prints the number of rows/columns, each column’s type, and sample values. This is a fast check for data types (numeric vs character), and for spotting obviously broken columns.

---

## 3. Basic numeric summary

```{r summary-stats}
# Summary for numeric variables (min/median/mean/max, NA counts implicitly visible)
safi %>%
  select(where(is.numeric)) %>%
  summary()
```

**Explanation (table output):** this summary provides quick descriptive statistics for numeric variables.
Use it to spot unrealistic values (e.g., negative counts) and to understand overall ranges before plotting.

---

## 4. Missing values overview

```{r missing-values}
# Top variables by number of missing values
library(tidyr)

safi %>%
  summarise(across(
    everything(),
    ~ sum(is.na(.))
  )) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) %>%
  arrange(desc(missing_count)) %>%
  head(10)
```

**Explanation (table output):** this table ranks variables by missingness.
It helps identify which columns may need cleaning, imputation, or removal depending on the analysis goal.

---

## 5. Exploratory charts

### 5.1 Household size distribution

```{r household-size-plot}
safi %>%
  filter(!is.na(no_membrs)) %>%
  ggplot(aes(x = no_membrs)) +
  geom_histogram(binwidth = 1, fill = "#1f78b4", color = "white", alpha = 0.9) +
  labs(
    title = "Distribution of Household Size",
    x = "Number of household members (no_membrs)",
    y = "Count of households"
  ) +
  theme_minimal()
```

**Explanation (chart):** a histogram shows the frequency of households by size.
This helps answer questions like “What is a typical household size?” and whether there are outliers.

---

### 5.2 Meals per day (food consumption indicator)

```{r meals-plot}
safi %>%
  filter(!is.na(no_meals)) %>%
  ggplot(aes(x = factor(no_meals), fill = factor(no_meals))) +
  geom_bar(color = "white", alpha = 0.9) +
  labs(
    title = "Number of Meals per Day",
    x = "Meals per day (no_meals)",
    y = "Number of households",
    fill = "Meals"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

**Explanation (chart):** a bar chart is appropriate for discrete values like meals per day.
It shows the most common meal counts (e.g., 2 or 3) and whether uncommon values appear.

---

### 5.3 Household size vs meals per day

```{r hh-meals-scatter}
safi %>%
  filter(!is.na(no_membrs), !is.na(no_meals)) %>%
  ggplot(aes(x = no_membrs, y = no_meals, color = factor(no_meals))) +
  geom_jitter(width = 0.2, height = 0.1, alpha = 0.7, size = 2) +
  labs(
    title = "Household Size vs. Meals per Day",
    x = "Household size (no_membrs)",
    y = "Meals per day (no_meals)",
    color = "Meals"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

**Explanation (chart):** this plot checks whether meals per day change with household size.
It is a visual check for association only and does not imply causality.

---

## 6. Short findings

- **Household size** varies across the dataset.
- **Meals per day** are concentrated in a few discrete categories.
- The **household size vs meals** plot shows no strong visual relationship.

---

## 7. Conclusion

This short EDA confirms the dataset loads correctly, provides a first look at distributions, missingness, and basic relationships.
