---
title: "Exploratory Data Analysis of SAFI Survey Data"
author: "Short Analytical Report"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## 1. Data description

This report provides a short exploratory data analysis (EDA) of the **SAFI survey dataset**.

**Unit of observation:** each row represents one household.

**Variables:** columns contain household characteristics, location identifiers, food consumption indicators, agricultural information, and related survey attributes.

**Goal of EDA:** quickly check the structure of the data, basic numeric distributions, missingness, and a few simple relationships using `dplyr` for selection/filtering and `ggplot2` for plots.

---

## 2. Data loading

```{r load-packages}
library(dplyr)
library(ggplot2)
library(readr)
```

```{r load-data}
# If SAFI_clean.csv is in the same folder as this .Rmd
safi <- read_csv("data/SAFI_clean.csv")
```

```{r data-structure}
# Data structure: column names, types, and example values
# This helps confirm that the dataset is loaded correctly.
glimpse(safi)
```

**Explanation (table output):** `glimpse()` prints the number of rows/columns, each column’s type, and sample values. This is a fast check for data types (numeric vs character), and for spotting obviously broken columns.

---

## 3. Basic numeric summary

```{r summary-stats}
# Summary for numeric variables (min/median/mean/max, NA counts implicitly visible)
safi %>%
  select(where(is.numeric)) %>%
  summary()
```

**Explanation (table output):** this summary provides quick descriptive statistics for numeric variables.
Use it to spot unrealistic values (e.g., negative counts) and to understand overall ranges before plotting.

---

## 4. Missing values overview

```{r missing-values}
# Top variables by number of missing values
# (Requires tidyr only for pivoting the output into a long table)
library(tidyr)

safi %>%
  summarise(across(
    everything(),
    ~ sum(is.na(.))
  )) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) %>%
  arrange(desc(missing_count)) %>%
  head(10)
```

**Explanation (table output):** this table ranks variables by missingness.
It helps identify which columns may need cleaning, imputation, or removal depending on the analysis goal.

---

## 5. Exploratory charts

### 5.1 Household size distribution

```{r household-size-plot}
# Identify the household size variable safely
# Common alternatives in SAFI-style data: household_members, members, hhsize
hh_size_var <- intersect(
  c("hh_members", "household_members", "members", "hhsize"),
  names(safi)
)[1]

stopifnot(!is.na(hh_size_var))

safi %>%
  filter(!is.na(.data[[hh_size_var]])) %>%
  ggplot(aes(x = .data[[hh_size_var]])) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of Household Size",
    x = "Number of household members",
    y = "Count of households"
  )
```

**Explanation (chart):** a histogram shows the frequency of households by size.
This helps answer questions like “What is a typical household size?” and whether there are outliers (very large households).

---

### 5.2 Meals per day (food consumption indicator)

```{r meals-plot}
# Bar chart: counts how many households reported each meals-per-day value
safi %>%
  filter(!is.na(meals)) %>%
  ggplot(aes(x = meals)) +
  geom_bar() +
  labs(
    title = "Number of Meals per Day",
    x = "Meals per day",
    y = "Number of households"
  )
```

**Explanation (chart):** a bar chart is appropriate for discrete values like meals per day.
It shows the most common meal counts (e.g., 2 or 3) and whether uncommon values appear.

---

### 5.3 Household size vs meals per day

```{r hh-meals-scatter}
# Use the detected household size variable
hh_size_var <- intersect(
  c("hh_members", "household_members", "members", "hhsize"),
  names(safi)
)[1]

stopifnot(!is.na(hh_size_var))

safi %>%
  filter(!is.na(.data[[hh_size_var]]), !is.na(meals)) %>%
  ggplot(aes(x = .data[[hh_size_var]], y = meals)) +
  geom_jitter(width = 0.2, height = 0.1, alpha = 0.6) +
  labs(
    title = "Household Size vs. Meals per Day",
    x = "Household size",
    y = "Meals per day"
  )
```

**Explanation (chart):** this plot checks whether meals per day change with household size.
If points show clear patterns (e.g., meals decreasing as household size increases), it may suggest constraints related to food access.
If the points are mixed without a trend, the relationship may be weak or require additional variables (e.g., income, location) to explain.

---

## 6. Short findings

- **Household size** varies across the dataset; the histogram indicates typical sizes and highlights any extreme values.
- **Meals per day** is mostly concentrated in a small set of categories, which makes it a simple proxy for food consumption conditions.
- The **household size vs meals** scatter does not necessarily imply causality; it is a quick check for visible association only.

---

## 7. Conclusion

This short EDA confirms the dataset loads correctly, provides a first look at distributions, missingness, and basic relationships.
Possible next steps include:

- comparing key indicators across villages/regions,
- building a simple food security index,
- modelling meals per day as a function of household and location factors.
